Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-11-29 inst/paper/main.tex
wip

\begin{abstract}
  \noindent Most dynamic languages allow users to turn text into code
  using various functions, often named \eval, with language-dependent
  semantics. The widespread use of these reflective functions hinders
  static analysis and prevents compilers from performing many
  optimizations. This paper aims to provide a better sense of why
  programmers use \eval.  Understanding how \eval is used in practice
  is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and
  application domain specific. In this paper, we focus on data science
  code written in R, and compare our results to previous work that
  analyzed web programming in JavaScript.  We conduct a large-scale
  study of a corpus of \CorpusAllCodeRnd lines of libraries and
  end-user code on which we performed dynamic analysis to confirm that
  \eval is indeed in widespread use. We found that \eval is more
  dangerous in some ways, and safer in others, than what was
  previously reported for JavaScript.
\end{abstract}

@@

The expressive power of \eval depends on choices made by the language
designers along the following two axes:
\begin{itemize}
\item {\bf Scoping:} What environment does the string passed to \eval
  execute in? JavaScript and R evaluate it in the current environment,
  thus exposing local variables and parameters to \eval and breaking
  the abstraction boundary of the enclosing function. Julia is more
  restrictive, it runs \eval at the ``top level'', in the global
  environment. JavaScript has a global \eval that behaves like Julia,
  it also has a ``strict mode'' in which \eval is allowed to access
  local variables, but fobidden from injecting new variables in the
  local environment. In Java, one can implement \eval but the
  expression will execute in new environment.\footnote{While \eval
  used the purview of interpreted languages, the advent of
  just-in-time compilation lifted this restriction. To implement \eval
  in Java requires wrapping its input in a static method of a new
  anonymous class, generating bytecode for that class, invoking the
  class loader to install that code, and finally reflectively calling
  the new method. }
\item {\bf Reflective API:} What other reflective operations are
  exposed to user code? JavaScript provides few reflective functions
  (other than enumeration of the properties of an object and access to
  properties with strings), \eval is thus limited to what can be
  expressed in JavaScript code. R, on the other hand, has a rich
  reflective interface. For example, R allows user code to walk the
  call stack and arbitrarily add and delete local variables in any
  environment. This reflective API means that there are few limits to
  what can do.
\end{itemize}

\noindent
These design choices impact our ability to reason about code when
reflection is used. A more expressive \eval translates directly into
additional restrictions for program analysis and transformation tools.
The reason Julia restricts \eval to the top level environment is to
preserve the compiler's ability to optimize
functions~\cite{oopsla18a}. What the Julia designers wanted was to
prevent \eval from modifying the type of local variables as this would
require recompilation of the method. Julia even adds a versoning
mechanism, called world age, to ensure that code added during an \eval
does not invalidate optimizations such as inlining~\cite{oopsla20a}.
More permissive languages such as R are much harder to deal with from
a compiler's perspective.

\medskip\noindent{\it Eval in R.} There are four reflective functions
that take abstract syntax trees (called expressions), parse them as
code, and evaluate the result. These functions are \c{eval},
\c{evalq}, \c{eval.parent}, and \c{local}. Taking an expression rather
than a string, may discourage some casual users as manipulating
expressions is slightly more involved.\footnote{The rationale for
\eval taking expressions is explained by the main use case for \eval,
namely extending the language by reinterpretation of promises; R gives
access to the code of a promise as an expression.} There are two ways
to obtain an expression. Given a string, the \c{parse()} function
retuns an unevaluated expression. The other way is to use the function
\c{substitute(e,envir)}, it yields the parse tree of the expression
\c{e} after performing substitutions defined by the bindings in the
\c{envir} parameter.

@@

\begin{lstlisting}
 eval(expr,
      envir = parent.frame(),
      enclos = if(is.list(envir) || is.pairlist(envir))
               parent.frame() else baseenv())
\end{lstlisting}

\noindent
This code is part of the definition of \eval, it specifies that the
function takes an \c{expr} which is the value to be evaluated,
\c{envir} is the environment in which evaluation happes, and
\c{enclos} is relevant when \c{envir} is a list or a data frame, then
it specifies where R looks for objects not found in \c{envir}. This
can be \c{NULL} (interpreted as the base package environment,
\c{baseenv()}) or an environment. Thus, the expression \c{expr}
evaluates in the environment \c{envir} and returns the computed value.
If \c{envir} is missing, then the default is \c{parent.frame()} (the
environment where the call to \eval was made). Values to be evaluated
can be of types \c{call} (function calls) or \c{expression} (arbitrary
expressions), \c{name} (when the symbol is looked up in the current
scope and its binding is evaluated), promise, or any of the basic
types such as vectors and environments (which are returned unchanged).

>>>>

\begin{abstract}
  \noindent Most dynamic languages allow users to turn text into code
  using various functions, often named \eval, with language-dependent
  semantics. The widespread use of these reflective functions hinders
  static analysis and prevents compilers from performing
  optimizations. This paper aims to provide a better sense of why
  programmers use \eval. Understanding why \eval is used in practice
  is key to finding ways to mitigate its negative impact. We have
  reasons to believe that reflective feature usage is language and
  application domain specific; we focus on data science code written
  in R, and compare our results to previous work that analyzed web
  programming in JavaScript. Dynamic analysis of a corpus of
  \CorpusAllCodeRnd lines of libraries and end-user code confirms that
  \eval is indeed in widespread use; R's \eval is more dangerous in
  some ways, and safer in others, than what was previously reported
  for JavaScript.
  %% TODO ::: Add a sentance about summarizing why users resort to
  %% eval -- Jan
\end{abstract}

@@

The expressive power of \eval depends on design decisions along two
axes:
\begin{itemize}
\item {\bf Scoping:} What environment does \eval executes its argument
  in? JavaScript and R evaluate it in the current environment, thus
  exposing local variables and parameters and breaking the caller's
  abstraction boundary. Julia is more restrictive, \eval runs at the
  ``top level'', in the global environment. JavaScript has a global
  \eval that behaves like Julia, and a ``strict mode'' in which \eval
  may access local variables, but fobidden from injecting new ones in
  the local environment. In Java, one can implement \eval with
  Julia-semantics.\footnote{While \eval used to be the purview of
  interpreted languages, just-in-time compilation lifted this
  restriction. \Eval takes its input, wraps it in a static method of a
  new anonymous class, generates bytecode for that class, invokes
  the class loader to install that code, and finally reflectively
  calls the method. }
\item {\bf Reflective API:} What other reflective operations are
  exposed to user code? JavaScript provides few reflective functions
  (other than enumeration of an object's properties and string-based
  indexing), the impact of \eval is thus limited. R, on the other
  hand, has a rich reflective API. For example, R allows user code to
  walk the call stack and arbitrarily add and delete local variables.
  This means that there are few limits to the side-effects of an
  \eval.
\end{itemize}

\noindent
These design choices impact our ability to reason about code when
reflection is used. A more expressive \eval translates directly into
additional restrictions for program analysis and transformation tools.
The reason Julia restricts \eval to the top level is to preserve the
compiler's ability to optimize code~\cite{oopsla18a}. In particular,
\eval had to be prevented from changing the type of local variables as
this would require recompilation of the method. Julia even adds a
versoning mechanism, called world age, to ensure that code added
during an \eval does not invalidate inlining
optimizations~\cite{oopsla20a}. More permissive languages such as R
are much harder to deal with from a compiler's perspective.

\medskip\noindent{\it Eval in R.} The reflective interface exposed by
R is somewhat complex. The core library exports four functions with
slightly different semantics, \c{eval}, \c{evalq}, \c{eval.parent},
and \c{local}. \Eval is the more general function, and, unlike in
JavaScript, it takes three arguments, an expression, its environment
and the enclosing environment. The following discussion simplify some
details which are not relevant to this paper. For the interested
reader we recommend the excellent book by Hadley
Wickham~\cite{hadley}. The definition of \eval starts as follows:

\begin{lstlisting}
 eval
   <- function(expr,
                envir = parent.frame(),
                enclos = if(is.list(envir)) parent.frame() else baseenv())
          { ...body... }
\end{lstlisting}
The parameters are \c{expr}, the value to be evaluated, \c{envir}, the
environment in which evaluation happens, and \c{enclos}, the
environment to look up objects not found in \c{envir} (by default
either the top level, \c{baseenv} or the environment of the caller).
The \c{expr} parameter can take values of different type, for our
purposes we focus on the most common, {\tt expression}s. It is easiest
to think of an \c{expression} as an abstract syntax tree as returned
by R's parser. The simplest way to create an \c{expression} is to call
\c{quote} passing some R expression:
\begin{lstlisting}
  > quote(a + b)
  # a + b
\end{lstlisting}
Here the return value is an abstract syntax tree. From a
string, use  the \c{parse} function:
\begin{lstlisting}
  > parse(text="a+b")
  # expression(a+b)
\end{lstlisting}

There are four reflective functions
that take abstract syntax trees (called expressions), parse them as
code, and evaluate the result. These functions are \c{eval},
\c{evalq}, \c{eval.parent}, and \c{local}. Taking an expression rather
than a string, may discourage some casual users as manipulating
expressions is slightly more involved.\footnote{The rationale for
\eval taking expressions is explained by the main use case for \eval,
namely extending the language by reinterpretation of promises; R gives
access to the code of a promise as an expression.} There are two ways
to obtain an expression. Given a string, the \c{parse()} function
retuns an unevaluated expression. The other way is to use the function
\c{substitute(e,envir)}, it yields the parse tree of the expression
\c{e} after performing substitutions defined by the bindings in the
\c{envir} parameter.

@@
